Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-06-18 WholePlotTrajectories.Rmd
relecture discussion

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   
    Understand the maintenance of forests biodiversity structure, functioning and dynamics is urgent to anticipate their fate in the global changing context.
    Communities diversity structure is assumed to rely on a constant regime of disturbance and peak at intermediate intensity.
    For tropical forests this intermediate disturbance hypothesis (IDH) remain debated as well as the completeness of forests resilience in all taxonomic and functoinal facets.

    To disentangle the ecological processes driving forests dynamics we examined the trajectories over 30 years after a gradient of logging and thinning disturbance, focusing on communities taxonomic and functional composition and diversity.
    Specifically we analysed the trajectories of communities taxonomic richness and evennes and functional composition, diversity and redundancy based on 7 key functional traits assessing species ecology.
    We highlighted a decoupling between functional and taxonomic trajectories, translating a fast recovery of communities functioning while their taxonomic recovery was slowed by the recruitment of infrequent species specific to mature forests.
    The IDH consistently predicted communities functional trajectories but poorly reflected the taxonomic ones, blurried by changes in the functional redundancy entailing competitive exclusion.
    Although communities consistently recovered their functioning after 30 years, initially infrequent species proved long to recover which impacted the taxonomic structure and functoinal redundancy over several decades.
    Our results acknowledged the need of decades-long recovery cycles to restore the resilience of the community and prevent species loss in the long term.

keywords: [Taxonomic and Functional Biodiversity, Neotropical Forests, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience]
journalinfo: Hal xxx
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::gitbook: default
  bookdown::word_document2: default
---

@@

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local climate, support the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The simultaneous increased demand in forests products and the substantial climatic changes currently heighten the pressure on remaining forests [@Gibson2011a; @Morales-Hidalgo2015], threatening the maintenance of communities structure, composition and functioning and their dynamics in space and time [@Anderson-Teixeira2013;@Sist2015].

In tropical forest, ecological communities are constantly re-shaped by natural disturbance events that change both abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and biotic interactions like competition among species [@Herault2018]. 
The cornerstone of tropical forests ecology is to understand the mechanisms and the determinants of ecosystems response to disturbance [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
These structural parameters were thereafter sucessfully modeled, giving important insights into the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009]. 
However the response of forests diversity in tree species remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014, @Liang2016] and would be most probably impacted by the changes following disturbance [@Baraloto2012a].

In the short-term disturbance demonstrated negligible or even positive impacts on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity at intermediate disturbance intensity [@Molino2001;@Kariuki2006a;@Berry2008a]. 
Still, validations of the IDH remain scarce in the long term and mainly rely on species richness analyses that give limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
More releveant monitoring would encompass communities composition, crucial for conservation issues, and complete diversity profiles integrating both species richness and evenness, to reveal the ecological rules structuring communities [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, the functional approach that accounts for species biological attributes and role in the ecosystem would be essential to understand the correlations between ecosystems biodiversity, functioning and environmental constraints [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
The functional trait-based approach focused on major traits related to species ecology and performance was sucessfully adopted [@Diaz2005;@Villeger2008a], for example in highlighting deterministic processes in tropical rainforests that foster fast growing species with efficient resources acquisition after disturbance [@Molino2001;@Haddad2008;@Ruger2009].
Communities then shifted from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
This shift translated into consistent trajectories for key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction life history traits (seed mass and maximum height) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].

A combination of taxonomic and functional approaches are essential to fully assess communities response to disturbance and differences in their respective trajectories are insightful of underlying mechanisms [@Lohbeck2015;@Guariguata2001].
Although taxonomic and functional diversity are complementary and sometimes decoupled, they are combined in the measure of communities functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016].
Functional redundancy is determinant of communities resilience as high redundancy, like in highly diverse tropical forests [@Bellwood2006], mitigates the impacts of species removal on ecosystem functioning [@Trenbath1999; @Elmqvist2003; @Diaz2005].

Grasp all facets of communities response to disturbance comes to examine the taxonomic and functional trajectories in diversity and composition.
These trajectories would highlight the ecological rules constraining or not communities dynamics towards the recovery of initial composition, diversity and functioning.
They would therefore clarify the tenants in the long term of the debated Intermediate Disturbance Hypothesis for tropical forests, a cruicial point for future adaptive conservation strategies [@Adler2007].
Here we monitored over 30 years the response of 75 ha of forests plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem biomass removed.
We made use of a large functional traits database browsing major leaf, stem and seed functional traits and species maximum height to draw the trajectories over time of communities taxonomic and functional composition and diversity. Specifically, we (i) questioned the coupling between taxonomic and functional response to disturbance and identified the underlying assembly processes, which allowed to (ii) clarify the validity of the IDH in the long term for tropical forest and (iii) question the completeness of communities recovery regarding their functional redundancy.

@@

The experiment is a network of twelve 6.25ha plots that underwent a gradient of three logging, thinning and fuelwood cutting treatments (Table \@ref(tab:Tab1)).
Disturbance treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Schmitt1989;@Blanc2009]. 
The disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance (ref to be found) measured with the BIOMASS R package [@Biomass2018].

@@

The study site corresponds to a tropical rainforest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical families. 
In the twelve experimental plots of the experiment, all trees above 10 cm DBH are mapped and measured annually since 1984. 
Trees are first identified during inventories with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names are given to 62 commercial or common species whereas more infrequent ones were identified under general identifiers only distinguiching trees and palm trees. 
From 2003 botanical campaigns were conducted every 5 to 6 years to identify all trees at the species level but identification practices still varied among plots and campaigns.

@@

Both communities taxonomic and functional diversity and composition proved resilient, following similar humped shaped trajectories returning towards initial values.
The resilience of communities functional structure, that is the most direct link between biodiversity and ecosystem functioning [@Diaz2005], meant a consistent recovery of ecosystem processes in the long term [@Guariguata2001].
The resilience of communities taxonomy meant the maintenance of their initial differences in composition and diversity, suggesting the existence of multiple stable equilibria as assumed for highly diverse and productive ecosystems [@Chase2003], and the dependency of recovery trajectories on the initial composition to be recovered that would the the asymptot of the humped shaped dynamics [@Hubbell1999; @Molino2001; @Anderson2007; @Baraloto2012a].

Although both communities taxonomic and functional characteristics proved resilient and followed similar humped shaped trajectories, the taxonomic recovery systematically lagged behind the functional one.
The decoupling between functional and taxonomic dynamics has already been observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015; @Guariguata2001].
According to the "vegetation quantity effect" [@Grime1998] functional trajectories rely on the pool of dominant species, which diversity and evenness increased after disturbance but rapidly recovered their initial functional structure.
Although the pool of dominant species recovered quickly, communities evenness remained high so infrequent species still missed to the taxonomic recovery.
Taxonomic recovery was then all the more longer that unrecovered species would be functionally redundant and probably underwent selective competition processes.

@@

The functional and taxonomic decoupling and the disprove of the IDh for communities taxonomic structure were explained were explained by persisting changes in communities functoinal redundancy that implied communities incomplete resilience [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Functional redundancy mitigates the loss of species on communities functional structure [@Carmona2016], and therefore reduces the impact on disturbance.
The functional redundancy itself however might be reduced or re-organized within the functional space  and alter communities resilience.
Although the functional redundancy considered in the whole communities functional space did not follow consistent trajectories after disturbance, the redundancy restricted to the functional space of the initial community clearly followed humped shaped trajectories.
This restricted redundancy first decreased, all the more so that the disturbance was intense, before returning towards the initial value but this had not recovered for any disturbed community after 30 years.
After disturbance communities differently occupied the functional space following traits trajectories and the shifts in functional composition, before recovering the initial functional space and its high redundancy.
This recovery phase was slow as it depend on infrequent, functionally redundant species specific to old-growth forests that rely upon different ecological processes, stochastic or deterministic competitive pressures limiting species functional similarity [@Chave2004;@Mayfield2010].

The unachieved recovery of the functional redundancy meant a decreased resilience of pre-disturbance communities and an increased of disturbance-specific communities (corresponding to lower richness and higher dominance of pioneer species).
After disturbance, chances were higher to observe long lasting or self-maintained compositional changes in favor of disturbance resistant species, lianas or epiphytes [@Haddad2008; @Burslem2000; @Martin2013].
The recovery remained incomplete until the recovery of the initial functional redundancy and resilience, and this specifically threatened the maintenance of species contingent to undisturbed forests and increased the risk of loosing cornerstone species and having unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].

@@

Our study defined the decoupled functional and taxonomic trajectories of tropical rainforests after disturbancethat demonstrated a rapid functional recovery but a slower and more variable taxonomic one.
Consistently with the IDH, functional trajectories were driven by deterministic processes favoring pioneers and light-demanding species after disturbance.
The following  functional shifts entailed a re-organization of communities functional redundancy in the functional space that proved long to recover and explained the more variable and longer taxonomic trajectories as shade tolerant species of old-growth forests faced functional redundancy and limiting similarity processes.
The resilience of tropical forests then proved long in the face of quite intense disturbance and did not preclude the settling of a persistent disturbanc-specific community [@Gourlet-Fleury2005].
Within the mechanisms of this resilience the recruitment processes proved central, in accordance with the IDH, and would deserve closer focus to specify their ecological drivers.
The disturbance range however stayed within the spectrum of selective logging, with a forest cover remaining all along the experiment, and the response mechanisms would probably be much different after harder disturbance.

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   
    Understand the maintenance of forests biodiversity structure, functioning and dynamics is urgent to anticipate their fate in the global changing context.
    Communities diversity are assumed to be shaped by constant regime of disturbance yielding a diversity peak at intermediate intensity, but for tropical forests this intermediate disturbance hypothesis (IDH) remains debated as well as the completeness their resilience in all taxonomic and functional facets after disturbance.

    To disentangle the ecological processes driving communities response to disturbance we analysed the taxonomic and functional trajectories following a logging and thinning disturbance gradient.
    Specifically we examined over 30 years the trajectories of communities taxonomic richness and evennes and functional composition, diversity and redundancy based on 7 leaf, stem and life history traits assessing species ecology.
    Trajectories highlighted communities taxonomic resilience, maintaining their initial differences in composition, and functional resilience through fast functional trajectory common to all communities in the functional space.
    The IDH consistently predicted communities functional trajectories according to the disturbance intensity, but poorly represented their taxonomic trajectories, blurried by the alteration of the functional redundancy and the recruitment of infrequent, shade tolerant species slowed by competitive exclusion processes.

    Although communities functioning had recovered after 30 years their taxonomic diversity and composition in infrequent species and their functional redundancy remained altered.
    these results acknowledged the need of decades-long recovery cycles to ensure communities complete resilience, and questioned communities maintenance after repeated disturbance.

keywords: [Taxonomic and Functional Biodiversity, Neotropical Forests, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience]
journalinfo: Hal xxx
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-ga-ni-za-tion}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
---

@@

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local climate, and the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The growing demand in forests products together with current climatic changes increases the pressure on remaining forests [@Gibson2011a; @Morales-Hidalgo2015]and threatens the maintenance and dynamics in space and time of communities structure, composition and functioning [@Anderson-Teixeira2013;@Sist2015].

In tropical forest, ecological communities are constantly re-shaped by natural disturbance events that change both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions like competition among species [@Herault2018]. 
The cornerstone of tropical forests ecology is to understand the processes and the determinants of ecosystems response to disturbance [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
These structural parameters were thereafter sucessfully modeled, giving important insights into the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009]. 
However the response of forests diversity in tree species remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014, @Liang2016] and would be most probably impacted by the changes induced by disturbance [@Baraloto2012a].

In the short-term disturbance demonstrated negligible or even positive impacts on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity at intermediate disturbance intensity [@Molino2001;@Kariuki2006a;@Berry2008a]. 
Still, validations of the IDH remain scarce in the long term and mainly rely on the analysis of species richness that gives limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
More complete analysis would besides encompass communities explicit composition that is crucial for conservation issues, and communities evenness to reveal the ecological rules structuring communities [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, a functional approach accoutning for species biological attributes and role in the ecosystem directly links ecosystems biodiversity, functioning and environmental constraints [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
The functional trait-based approach focused on major traits related to species ecology and performance was sucessfully adopted [@Diaz2005;@Villeger2008a], for example to reveal in tropical rainforests the deterministic processes that foster after disturbance the fast growing species with efficient resources acquisition [@Molino2001;@Haddad2008;@Ruger2009].
This basic deterministic process entails a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
This shift is translated into consistent trajectories of key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction life history traits (seed mass and maximum height) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Besides communities functional composition and diversity, a key aspect of communties response to disturbance is the changes in functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016].
High functional redundancy, as in the very diverse tropical forests [@Bellwood2006], mitigates the impacts of species removal on ecosystem functioning and determines communities resilience [@Trenbath1999; @Elmqvist2003; @Diaz2005].
High functional redundancy also increases the functional overlap among species and would foster neutral stochastic processes [@Gravel2006].

To grasp all facets of communities response to disturbance we examined here the taxonomic and functional trajectories in diversity and composition [@Lohbeck2015;@Guariguata2001].
These trajectories would highlight the recovery of communities initial composition, diversity and functioning and the underlying ecological processes.
They would clarify the tenants in the long term of the Intermediate Disturbance Hypothesis debated in tropical forests, and provide indications for future adaptive conservation strategies [@Adler2007].
We monitored over 30 years the response of 75 ha of neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem biomass removed.
We made use of a large functional traits database browsing major leaf, stem and seed functional traits and species maximum height to draw the trajectories over time of communities taxonomic and functional composition and diversity. Specifically, we (i) questioned the coupling between taxonomic and functional response to disturbance and identified the underlying assembly processes, (ii) clarified the validity of the IDH in the long term for tropical forest and (iii) questioned the completeness of communities recovery regarding communities functional redundancy.

@@

The experiment is a network of twelve 6.25ha plots that underwent a gradient of three logging, thinning and fuelwood cutting treatments (Table \@ref(tab:Tab1)) attributed according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Schmitt1989;@Blanc2009]. 
The disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance (ref to be found) measured with the BIOMASS R package [@Biomass2018].

@@

The study site corresponds to a tropical rainforest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical families. 
In the twelve experimental plots of the experiment, all trees above 10 cm DBH are mapped and measured annually since 1984. 
Trees are first identified during inventories with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names are given to 62 commercial or common species whereas more infrequent ones were identified under general identifiers only distinguishing trees and palm trees. 
From 2003 botanical campaigns were conducted every 5 to 6 years to identify all trees at the species level but identification practices still varied among plots and campaigns.

@@

Both communities taxonomic and functional diversities and composition proved resilient, following similar humped shaped trajectories returning towards initial values.
Communities functional structure is the most direct link between biodiversity and ecosystem functioning [@Diaz2005] and its resilience meant the recovery of ecosystem processes in the long term [@Guariguata2001].
The resilience of communities taxonomy in turn meant the maintenance of their initial differences in species composition and diversity.
The maintenance of initial differences suggested the existence of multiple stable equilibria as assumed for highly diverse and productive ecosystems [@Chase2003] and the dependency of recovery trajectories on the initial composition to be recovered [@Hubbell1999; @Molino2001; @Anderson2007; @Baraloto2012a].

Although communities taxonomic and functional trajectories were similar the taxonomic recovery lagged behind the functional one, confirming a decoupling between functional and taxonomic dynamics already observed for grasslands [@Tilman1997; @Mouillot2011] and tropical forests [@Lohbeck2015; @Guariguata2001].
According to the "vegetation quantity effect" [@Grime1998] the functional trajectories rely on the pool of dominant species, which diversity and evenness increased and rapidly recovered after disturbance.
At the same time communities taxonomic evenness remained high, revealing the unachieved recovery of infrequent species.

@@

The functional and taxonomic decoupling and the disprove of the IDH for communities taxonomic structure were explained by persisting rearranging of communities functional redundancy that implied an incomplete resilience of communities [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Although the functional redundancy considered in the functional space of the whole communities did not display consistent trajectories after disturbance, the redundancy restricted to the functional space of the initial community clearly followed humped shaped trajectories.
After disturbance communities differently occupied the functional space, following the observed shifts in traits trajectories and functional composition, and the redundancy in the initial functional space decreased according to the disturbance intensity.
After 30 years the recovery of the initial redundancy structure was consistent but remained unachieved.
It depended on infrequent species specific to old-growth forests that were functionally and therefore underwent competitive exclusion limiting functional similarity [@Chave2004;@Mayfield2010].
The persistent alteration of functional redundancy meant a lower resilience of pre-disturbance communities and an higher resilience disturbance-specific communities (corresponding to lower richness and higher dominance of pioneer species) so chances to observe self-maintained compositional changes in favor of disturbance resistant species, lianas or epiphytes were increased after disturbance [@Haddad2008; @Burslem2000; @Martin2013].
Communities resilience depended on the functional redundancy structure that proved long to recover.
This decades-long recovery specifically increased the risks of loosing cornerstone specie, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].

@@

Our study defined the decoupled functional and taxonomic trajectories of tropical rainforests after disturbance demonstrating a rapid functional recovery but a slower and more variable taxonomic one.
Consistently with the IDH, functional trajectories were driven after disturbance by deterministic processes favoring pioneers and light-demanding species.
The following functional shifts entailed a re-organization of communities functional redundancy that proved long to recover and explained the more variable and longer taxonomic trajectories as shade tolerant species of old-growth forests faced functional redundancy and limiting similarity processes.
The resilience of tropical forests then proved long in the face of quite intense disturbance and did not preclude the settling of a persistent disturbance-specific community [@Gourlet-Fleury2005].
As the trajectories highlighted the recruitment processes proved central for communities response to disturbance and closer focus on demographical drivers of communities response would clarify the fate of the future forests.
The disturbance range however stayed within the spectrum of selective logging, with a forest cover remaining all along the experiment, and the response mechanisms would probably be much different after harder disturbance.
